---
title: "Creating maps with ggplot2+sf and ggmap"
author: "Mel Moreno"
output:
  html_document:
    fig_caption: no
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '2'
---


```{r global_options, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
    results = "hide",
    warning = FALSE, 
    message = FALSE,
    cache = TRUE)
## This is just to "pre-load" all packages, and prevent the startup
## messages to show
## devtools::install_github("tidyverse/ggplot2")
## devtools::install_github("3wen/legendMap")
library(devtools)
library(ggmap)
library(ggplot2)
library(ggthemes)
library(legendMap)
library(maps)
library(maptools)
library(rgdal)
library(rgeos)
library(rworldmap)
library(rworldxtra)
library(sf)
library(sp)
library(tidyverse)
library(tools)
library(dplyr)
library(cowplot)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
library (viridis)

```

# Plotting layers, and additional features

##Introduction

In our other maps, we were creating general maps with very little landscape information. Some maps will require water or terrain features to enhance your data. We will be using the spatial data found in this link http://www.mapcruzin.com/free-united-states-shapefiles/free-florida-arcgis-maps-shapefiles.htm, so you can follow along. You can download some of the packages and unzip them to create the layers on the next map.

Sometimes, however, you will be provided better or more accurate shapefile data, but we will use this public spatial data to become familiar with adding spatial data layers.

The packages used for this map are listed for convenience.



```{r featmap_packages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(rgdal)
library(rgeos)
library(maps)
library(maptools)
library(rworldmap)
library(rworldxtra)
library(devtools)
library(legendMap)
library(tidyverse)
library(ggmap)
```

## Data and basic plot

The package `rgdal` is necessary to allow us to read the shapefile spatial data. In this tutorial we will show you two different ways to retrieve shapefiles spatial data. Some instances you will have data provided to you, so you will be able to download that spatial data directly from a local network. In this case, we will gather our spatial data from an open source website. You can follow along to download spatial data here[fgdl](http://www.fgdl.org), but you can continue using the next code for downloading directly from a URL. 

The function `readOGR` will require a `dsn`, which is what folder your spatial data is located, and also a `layer`, which is the actual spatial data file, without the .shp at the end. 

Credit:Fgdl.org

```{r pkg-rworldmap}
library("rworldmap")
library("rworldxtra")
```
```{r world}
world <- getMap(resolution = "high")
class(world)
```

The world map is available as a `SpatialPolygonsDataFrame`; we thus
convert it to a simple feature using `st_as_sf` (from package `sf`):

```{r world-sf}
world <- st_as_sf(world)
```

```{r featmap_datamanual, eval=FALSE }

library(rgdal)

flstparks<-readOGR(dsn="E:/Basille Lab/LayersShapefiles/stpark_dec15",layer="stpark_dec15")

flmjwater<-readOGR(dsn="E:/Basille Lab/LayersShapefiles/mjwaterbnd",layer="mjwaterbnd")

flparksandrec<-readOGR(dsn="E:/Basille Lab/LayersShapefiles/gc_parksbnd_jul16",layer="gc_parksbnd_jul16")

flhighway<-readOGR(dsn="E:/Basille Lab/LayersShapefiles/interstates",layer="interstates")
```

Other instances will require you to retrieve spatial data on the internet to complete your map. We can batch of our links in the `URLs`, and the rest of the function is a generic code. The code can be used to retrieve multiple spatial data from one source, in our case it is www.fdgl.org.  

```{r featmap_dataURL, eval=FALSE}
require(knitr)
opts_knit$set(root.dir = normalizePath('../'))

shp <- c("mjwaterbnd", "gc_parksbnd_jul16", "stpark_dec15", "rci_off_apr17")
ftp <- "ftp://ftp.fgdl.org/pub/state/"

wd <- getwd()
setwd(tempdir())
for (k in shp) {
    download.file(paste0(ftp, k, ".zip"), paste0(k, ".zip"))
    unzip(paste0(k, ".zip"))
    assign(k, st_read(paste0(k, ".shp"), quiet = TRUE))
}
setwd(wd)

```

To make sure all of our layers are compatible which each other, we will be tranforming every layer into the EPSG 4326, which is datum WGS 84. Our layers will then be projected accurately on our map because they will be in the same coordinate projection. The function `st_transform` is part of the `sf` package and will convert the spatial data into an `sf` object as well as convert the coordinate projection to EPSG 4326. This step is necessary whether you retrieve the data from a local source, or an open internet source. 

```{r featmap_transform, eval=FALSE}
flstparks2 <- st_transform(stpark_dec15, crs = 4326)
flmjwater2 <- st_transform(mjwaterbnd, crs = 4326)
flparksandrec2 <- st_transform(gc_parksbnd_jul16, crs = 4326)
flhighway2 <- st_transform(rci_off_apr17, crs = 4326)
```


## Creating Cities 

We would like to add the main Florida cities to our map, as opposed to all of the cities in Florida because it might make it look very cluttered. You can use this general code for any city in the United States. You could change the region to the state name, which is lowercase. The function `geocode` requires the package `ggmap` .

```{r featmap_cities}
library(ggmap)

flcities <- data.frame(State=rep("florida",7), 
                         City=c("Miami", "Fort Lauderdale", "Orlando", "Gainesville", "Tallahassee", "Tampa","Jacksonville"))

flcities<-cbind(geocode(as.character(flcities$City)),flcities)

```

Now that we have specified our cities, we can plot them unto our Florida map. We will use `geom_text` to add the text layer unto our map. The funciton `geom_text` requires that we add an aesthetic variable, `aes()` that will inform the function what the x and y axis are and what our label is. We specified the `City` in our `flcities` command in the previous code post. The function will also require were the data is coming from, the size you wish to set the text (default it is size 3), and we also nudged the text to be slighty over the coordinates where the cities are found. The argument `check_overlap` will remove any text that crosses over each other, and we lastly chose to make our text bold. 

```{r}
ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") + 
    geom_text(aes(x=lon, y=lat, label = City), data = flcities, size = 3.9, nudge_y =.18, col = "black", check_overlap = TRUE,fontface = "bold") +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) 
```


## Annotated text

We used previously `geom_text` to add the state names in the Coordinates Points Map, but we can also use `annotate` to manually add text to our map. In this case we will be adding a text for the Gulf of Mexico. We specify that we will be using a geom_text in the `geom=` argument. We will also specify where we would like the text on the map using the x and y. Our label is the actual text `Gulf of Mexico`. We also can describe the color and the sizes of the text.

```{r featmap_annotext}
ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) +
    annotate(geom = "text", x = -85.5, y = 27.5, label = "Gulf of Mexico", 
        color = "grey22", size = 4.5)
```


## Final Map

The final map will include all the layers we retrieved from previous code. Some instances you will see `color`, `colour`, and `col` being used, but they all mean the same code. You can use `col` and it will apply the same effect as `color`. Before we applied `coord_sf` at the beginning of our `ggplot` code, but when applying all of the layers, it will work after all of the layers specified. That is why you will see it after the `geom_sf` `flhighway2` line code. You can see that `geom_sf` can be used on a variety of spatial data.

```{r coord_getState}
library("tigris")
library("sp")
library("sf")

st <- states(resolution = "20m", class="sf")
st <- st_transform(st, crs=4326)
head(st)

```


```{r featmap_final}
ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") +
    #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
    #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
    #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
    #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
    geom_sf(data = st, fill = NA, colour = "black") + 
    geom_text(aes(x=lon, y=lat, label = City), data = flcities, size = 3.9, inherit.aes = TRUE, nudge_y =.18, col = "black", check_overlap = TRUE,fontface = "bold") +
    annotate(geom="text", x=-85.50, y=27.50, label= "Gulf of Mexico", color="grey22",size=4.5) +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) +
    ggtitle ("Florida Natural Resources Map") + 
    xlab("Longitude")+ ylab("Latitude") +
    theme_classic() + 
    theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA)) +
    scale_bar(lon = -87.00, lat =24.50,distance_lon = 100, distance_lat = 15, distance_legend = 50, dist_unit = "km", arrow_length=50, arrow_distance=65, arrow_north_size =5, orientation= TRUE,legend_size=2.5) 
```



# Print Feature

You can use the `print` function to place multiple maps in one map graphic. This can be a useful and easy way to enhance the information you need to display.  


```{r printmap_mapsdata, include=FALSE}


  rmarkplot<-
  
  ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") +
    #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
    #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
    #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
    #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
    geom_sf(data = st, fill = NA, colour = "black") + 
    geom_text(aes(x=lon, y=lat, label = City), data = flcities, size = 3.0, inherit.aes = TRUE, nudge_y =.18, col = "black", check_overlap = TRUE,fontface = "bold") +
    annotate(geom="text", x=-85.50, y=27.50, label= "Gulf of Mexico", color="grey22",size=4.5) +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) +
    xlab("Longitude")+ ylab("Latitude") +
    theme_classic() + 
    theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA))

rmarkplot2<-  
  
  ggplot(data=world) + 
  geom_sf(fill="antiquewhite1") +
  #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
  #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
  #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
  #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.63, 26.9), expand=TRUE ) +
    ggtitle ("Location A") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))

rmarkplot3<- 
  ggplot(data=world)+
  geom_sf(fill="antiquewhite1") +
  #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
  #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
  #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
  #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.62, 26.35), expand=TRUE ) + 
  ggtitle ("Location B") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))
  
  

```


To use the print function you would need to first make view port that specify the max width and height of each map, and the minimum x and y coordinates where it map can be located. The `just=` will make a position on how the secondary maps will be displayed. I wanted the secondary maps to be displayed on the bottom left of the frame, so that will create a guide on where they will first appear. 

```{r printmap_vp}
vp <- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.45, just = c("left","bottom")) 
vp1<- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.15, just = c("left","bottom")) 
```

When the maps are completed, and the `viewport`s are specified, you can now print the secondary maps on top of your main map. This will take some tweeking, since it's not measured in coordinates but rather with R Studios own measurements. 

```{r printmap_finalmap}

print(rmarkplot)
print(rmarkplot2, vp=vp)
print(rmarkplot3, vp=vp1)


```


You can also decide to save this printed map by running all the code at once. It will save it in the format you prefer. We used it as a png. but pdf. is also available. 

```{r printmap_saving, eval=FALSE}
png("PrintMapping.png",width = 7, height = 7, units = "in", res = 300)
print(rmarkplot)
print(rmarkplot2, vp=vp)
print(rmarkplot3, vp=vp1)
dev.off()
```


# Multiple Maps in one window

## Introduction

Having a way to emphasize locations on a map can also be a graphic that would be useful for biologists. The next steps will be to create blank sqaure plots that will make your map visualization interesting and accurate. In our map, we will zooming into two different locations in Florida, for a close up of the area using the map we previously built. 

First we will start with the overall packages needed to build this 

Credit to Ilya Kashnitsky (https://ikashnitsky.github.io/2017/align-six-maps/).

```{r listplot_packages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(cowplot)
library(ggthemes)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
```


## Map Placement Function 
 The function below will allow us to grab plots that we have previously made, with `ggplotGrob` , and create arrows pointing to the maps. 
 
First we need to list our plots, normally in the order you would like them to appear. We have create 

```{r listplots_function}

list.plots=list(rmarkplot,rmarkplot2, rmarkplot3)
align_three_plots <- function(list.plots, family = "")
  {
  
gg <- ggplot()+
 
  coord_equal(xlim = c(0, 80), ylim = c(0, 51), expand=TRUE) +
  annotation_custom(ggplotGrob(rmarkplot), xmin = 0.0, xmax =57, ymin = 0.0, ymax = 50) +
  annotation_custom(ggplotGrob(rmarkplot2),xmin = 58, xmax = 80, ymin = 25, ymax = 51) +
  annotation_custom(ggplotGrob(rmarkplot3), xmin = 58, xmax = 80, ymin = 0.0, ymax = 29) +
  labs(x = NULL, y = NULL) +
  theme_void()



#DF with the coordinates of the 2 arrows
df.arrows <- data.frame(id=1:2,
                        x=c(45.6,45.4),
                        y=c(21.7,19.8),
                        xend=c(58,58),
                        yend=c(35,12))


gg <- gg +
  geom_segment(data = df.arrows %>% filter(id==1),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm"))) +
  geom_segment(data = df.arrows %>% filter(id==2),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm")))
}

```


## Creating a Simple Empty Square Plot

The next code will allow up to make a simple empty square plot, that can be cloned so we can place our maps inside. We are only using 

```{r listplot_blankplot}

library("ggthemes")

p <- 
  ggplot()+
  theme_map() +
  theme(panel.border = element_rect(color = "black", size = 0.5, fill = NA),aspect.ratio = 1)

```

## Cloning and Final Map 

```{r}

library("tidyverse")

plots <- mget(rep("p", 3))

three<- align_three_plots(plots)

plot(three)
```

## Maps with legend inbetween


```{r}

LegendMap<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  scale_fill_viridis(option="plasma", discrete=TRUE)

Map1<- 
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-10.00, 40.00), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

plot(Map1)

Map2<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-100.00,-60.00 ), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

plot(Map2)
  

```





```{r}
# Now grabbing the legends from LegendMap map and plotting them in the same window, the x11 is the size of the window I want, the NULL in the plot_grid is to make an empty space where I can later draw the legend with draw_grob

legend <- get_legend(LegendMap+ theme(legend.position="left") )


prow <- plot_grid( Map1 + theme(legend.position="none"), 
                   NULL, Map2 + theme(legend.position="none"),
                   align = 'vh',
                   #labels = c("A","", "B" ),
                   hjust = -1,
                   nrow = 1,
                   rel_widths = c(1, .50, 1)
)
x11(width=12, height=6, pointsize=10)

prow + draw_grob(legend, 2/4,1/8, 0.003) 


```









# Helpful Resources 

##PDFs for packages

-ggplot2 
https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf
http://ggplot2.tidyverse.org/reference/stat_ecdf.html

-sf
https://cran.r-project.org/web/packages/sf/sf.pdf

-sp
https://cran.r-project.org/web/packages/sp/sp.pdf

-CRAN
https://cran.r-project.org/

-Viridis Color Palettes
https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

-Scale Bar
https://rdrr.io/github/3wen/legendMap/man/scale_bar.html

##Links and various tutorials

-Global Choropleth, world interactive, rotating map

http://ellisp.github.io/blog/2017/06/04/military-gdp


-Ggplotly

http://www.alexejgossmann.com/salaries_by_school_plotly_viz/


-R studio colors

http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf


-Plotting side by side

https://ikashnitsky.github.io/2017/align-six-maps/


-Leaflet

https://cengel.github.io/rspatial/4_Mapping.nb.html

Points with gganimate
http://blog.cultureofinsight.com/2017/06/building-dot-density-maps-with-uk-census-data-in-r/


-Symobls and Points
http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r


-Themes
http://ggplot2.tidyverse.org/reference/theme.html


-Two graphs with same legend

https://github.com/tidyverse/ggplot2/wiki/share-a-legend-between-two-ggplot2-graphs


- Downloading Multiple Zipped files
https://www.r-bloggers.com/batch-downloading-zipped-shapefiles-with-r/

