---
title: "Positioning maps made with ggplot2+sf"
author: "Mel Moreno and Mathieu Basille"
output:
  html_document:
    fig_caption: no
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r global_options, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
    results = "hide",
    warning = FALSE, 
    message = FALSE,
    cache = TRUE)
## This is just to "pre-load" all packages, and prevent the startup
## messages to show
## devtools::install_github("tidyverse/ggplot2")
## devtools::install_github("3wen/legendMap")
library(devtools)
library(ggmap)
library(ggplot2)
library(ggthemes)
library(legendMap)
library(maps)
library(maptools)
library(rgdal)
library(rgeos)
library(rworldmap)
library(rworldxtra)
library(sf)
library(sp)
library(tidyverse)
library(tools)
library(dplyr)
library(cowplot)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
library (viridis)

```

# Map placement




```{r featmap_packages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(rgdal)
library(rgeos)
library(maps)
library(maptools)
library(rworldmap)
library(rworldxtra)
library(devtools)
library(legendMap)
library(tidyverse)
library(ggmap)
```


```{r pkg-rworldmap, include= FALSE}
library("rworldmap")
library("rworldxtra")
library("ggmap")
library("tigris")
library("sp")
library("sf")

world <- getMap(resolution = "high")
world <- st_as_sf(world)

flcities <- data.frame(State=rep("florida",7), 
                         City=c("Miami", "Fort Lauderdale", "Orlando", "Gainesville", "Tallahassee", "Tampa","Jacksonville"))

flcities<-cbind(geocode(as.character(flcities$City)),flcities)
st <- states(resolution = "20m", class="sf")
st <- st_transform(st, crs=4326)

```


# Print Feature

You can use the `print` function to place multiple maps in one map graphic. This can be a useful and easy way to enhance the information you need to display.  


```{r printmap_mapsdata, include=FALSE}

  rmarkplot<-
  
  ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") +
    #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
    #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
    #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
    #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
    geom_sf(data = st, fill = NA, colour = "black") + 
    geom_text(aes(x=lon, y=lat, label = City), data = flcities, size = 3.0, inherit.aes = TRUE, nudge_y =.18, col = "black", check_overlap = TRUE,fontface = "bold") +
    annotate(geom="text", x=-85.50, y=27.50, label= "Gulf of Mexico", color="grey22",size=4.5) +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) +
    xlab("Longitude")+ ylab("Latitude") +
    theme_classic() + 
    theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA))

rmarkplot2<-  
  
  ggplot(data=world) + 
  geom_sf(fill="antiquewhite1") +
  #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
  #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
  #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
  #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.63, 26.9), expand=TRUE ) +
    ggtitle ("Location A") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))

rmarkplot3<- 
  ggplot(data=world)+
  geom_sf(fill="antiquewhite1") +
  #geom_sf(data = flstparks2, colour = "forestgreen", fill = "forestgreen") +
  #geom_sf(data = flparksandrec2, fill = "darkgreen", colour = "darkgreen", alpha = 0.8) +
  #geom_sf(data = flmjwater2, fill = "deepskyblue", colour = "deepskyblue", alpha = 0.5) +
  #geom_sf(data = flhighway2, col = "grey50", size = 0.65) +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.62, 26.35), expand=TRUE ) + 
  ggtitle ("Location B") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))
  
  

```


To use the print function you would need to first make view port that specify the max width and height of each map, and the minimum x and y coordinates where it map can be located. The `just=` will make a position on how the secondary maps will be displayed. I wanted the secondary maps to be displayed on the bottom left of the frame, so that will create a guide on where they will first appear. 

```{r printmap_vp}
vp <- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.45, just = c("left","bottom")) 
vp1<- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.15, just = c("left","bottom")) 
```

When the maps are completed, and the `viewport`s are specified, you can now print the secondary maps on top of your main map. This will take some tweeking, since it's not measured in coordinates but rather with R Studios own measurements. 

```{r printmap_finalmap}

print(rmarkplot)
print(rmarkplot2, vp=vp)
print(rmarkplot3, vp=vp1)


```


You can also decide to save this printed map by running all the code at once. It will save it in the format you prefer. We used it as a png. but pdf. is also available. 

```{r printmap_saving, eval=FALSE}
png("PrintMapping.png",width = 7, height = 7, units = "in", res = 300)
print(rmarkplot)
print(rmarkplot2, vp=vp)
print(rmarkplot3, vp=vp1)
dev.off()
```


# Multiple Maps in one window

## Introduction

Having a way to emphasize locations on a map can also be a graphic that would be useful for biologists. The next steps will be to create blank sqaure plots that will make your map visualization interesting and accurate. In our map, we will zooming into two different locations in Florida, for a close up of the area using the map we previously built. 

First we will start with the overall packages needed to build this 

Credit to Ilya Kashnitsky (https://ikashnitsky.github.io/2017/align-six-maps/).

```{r listplot_packages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(cowplot)
library(ggthemes)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
```


## Map Placement Function 
 The function below will allow us to grab plots that we have previously made, with `ggplotGrob` , and create arrows pointing to the maps. 
 
First we need to list our plots, normally in the order you would like them to appear. We have create 

```{r listplots_function}

list.plots=list(rmarkplot,rmarkplot2, rmarkplot3)
align_three_plots <- function(list.plots, family = "")
  {
  
gg <- ggplot()+
 
  coord_equal(xlim = c(0, 80), ylim = c(0, 51), expand=TRUE) +
  annotation_custom(ggplotGrob(rmarkplot), xmin = 0.0, xmax =57, ymin = 0.0, ymax = 50) +
  annotation_custom(ggplotGrob(rmarkplot2),xmin = 58, xmax = 80, ymin = 25, ymax = 51) +
  annotation_custom(ggplotGrob(rmarkplot3), xmin = 58, xmax = 80, ymin = 0.0, ymax = 29) +
  labs(x = NULL, y = NULL) +
  theme_void()



#DF with the coordinates of the 2 arrows
df.arrows <- data.frame(id=1:2,
                        x=c(45.6,45.4),
                        y=c(21.7,19.8),
                        xend=c(58,58),
                        yend=c(35,12))


gg <- gg +
  geom_segment(data = df.arrows %>% filter(id==1),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm"))) +
  geom_segment(data = df.arrows %>% filter(id==2),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm")))
}

```


## Creating a Simple Empty Square Plot

The next code will allow up to make a simple empty square plot, that can be cloned so we can place our maps inside. We are only using 

```{r listplot_blankplot}

library("ggthemes")

p <- 
  ggplot()+
  theme_map() +
  theme(panel.border = element_rect(color = "black", size = 0.5, fill = NA),aspect.ratio = 1)

```

## Cloning and Final Map 

```{r}

library("tidyverse")

plots <- mget(rep("p", 3))

three<- align_three_plots(plots)

plot(three)
```

## Maps with legend inbetween


```{r}

LegendMap<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  scale_fill_viridis(option="plasma", discrete=TRUE)

Map1<- 
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-10.00, 40.00), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

plot(Map1)

Map2<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-100.00,-60.00 ), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

plot(Map2)
  

```





```{r}
# Now grabbing the legends from LegendMap map and plotting them in the same window, the x11 is the size of the window I want, the NULL in the plot_grid is to make an empty space where I can later draw the legend with draw_grob

legend <- get_legend(LegendMap+ theme(legend.position="left") )


prow <- plot_grid( Map1 + theme(legend.position="none"), 
                   NULL, Map2 + theme(legend.position="none"),
                   align = 'vh',
                   #labels = c("A","", "B" ),
                   hjust = -1,
                   nrow = 1,
                   rel_widths = c(1, .50, 1)
)
x11(width=12, height=6, pointsize=10)

prow + draw_grob(legend, 2/4,1/8, 0.003) 


```









# Helpful Resources 

##PDFs for packages

-ggplot2 
https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf
http://ggplot2.tidyverse.org/reference/stat_ecdf.html

-sf
https://cran.r-project.org/web/packages/sf/sf.pdf

-sp
https://cran.r-project.org/web/packages/sp/sp.pdf

-CRAN
https://cran.r-project.org/

-Viridis Color Palettes
https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

-Scale Bar
https://rdrr.io/github/3wen/legendMap/man/scale_bar.html

##Links and various tutorials

-Global Choropleth, world interactive, rotating map

http://ellisp.github.io/blog/2017/06/04/military-gdp


-Ggplotly

http://www.alexejgossmann.com/salaries_by_school_plotly_viz/


-R studio colors

http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf


-Plotting side by side

https://ikashnitsky.github.io/2017/align-six-maps/


-Leaflet

https://cengel.github.io/rspatial/4_Mapping.nb.html

Points with gganimate
http://blog.cultureofinsight.com/2017/06/building-dot-density-maps-with-uk-census-data-in-r/


-Symobls and Points
http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r


-Themes
http://ggplot2.tidyverse.org/reference/theme.html


-Two graphs with same legend

https://github.com/tidyverse/ggplot2/wiki/share-a-legend-between-two-ggplot2-graphs


- Downloading Multiple Zipped files
https://www.r-bloggers.com/batch-downloading-zipped-shapefiles-with-r/

