---
title: "Positioning maps made with ggplot2+sf"
author: "Mel Moreno and Mathieu Basille"
output:
  html_document:
    fig_caption: no
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r global_options, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
    results = "hide",
    warning = FALSE, 
    message = FALSE,
    cache = TRUE)
## This is just to "pre-load" all packages, and prevent the startup
## messages to show
## devtools::install_github("tidyverse/ggplot2")
## devtools::install_github("3wen/legendMap")
library(devtools)
library(ggmap)
library(ggplot2)
library(ggthemes)
library(legendMap)
library(maps)
library(maptools)
library(rgdal)
library(rgeos)
library(rworldmap)
library(rworldxtra)
library(sf)
library(sp)
library(tidyverse)
library(tools)
library(dplyr)
library(cowplot)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
library (viridis)

```

# Map placement
Depending on the visual information that needs to be displayed, maps and their corresponding data might need to be arrange to create easy to read graphical representations. This can be a very useful tools for publications or presentations. Sometimes adding a slightly different placement to a map, will make the information portrayed more aesthetically appealing. 


```{r mappackages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(rgdal)
library(rgeos)
library(maps)
library(maptools)
library(rworldmap)
library(rworldxtra)
library(devtools)
library(legendMap)
library(tidyverse)
library(ggmap)
```


```{r pkg-rworldmap, include= FALSE}
library("rworldmap")
library("rworldxtra")
library("ggmap")
library("tigris")
library("sp")
library("sf")

world <- getMap(resolution = "high")
world <- st_as_sf(world)

flcities <- data.frame(State=rep("florida",7), 
                         City=c("Miami", "Fort Lauderdale", "Orlando", "Gainesville", "Tallahassee", "Tampa","Jacksonville"))

flcities<-cbind(geocode(as.character(flcities$City)),flcities)
st <- states(resolution = "20m", class="sf")
st <- st_transform(st, crs=4326)

sites <- data.frame(longitude = c(-80.144005, -80.109), latitude = c(26.479005, 
    26.83))

sites <- st_as_sf(sites, coords = c("longitude", "latitude"), 
    crs = 4326, agr = "constant")

```


# Print Feature

The `print` command can be used place multiple maps in one map graphic. Essentially the `print` command This can be a useful and easy way to enhance the information you need to display.  


```{r print_mapsdata, include=FALSE}

  rmarkplot<-
   ggplot(data = world) +
    geom_sf(fill = "antiquewhite1") +
    geom_sf(data = sites, size = 4, shape = 23, fill = "darkred") +
    geom_sf(data = st, fill = NA, colour = "black") + 
    geom_text(aes(x=lon, y=lat, label = City), data = flcities, size = 3.0, inherit.aes = TRUE, nudge_y =.18, col = "black", check_overlap = TRUE,fontface = "bold") +
    annotate(geom="text", x=-85.50, y=27.50, label= "Gulf of Mexico", color="grey22",size=4.5) +
    coord_sf(xlim = c(-87.35, -79.5), ylim = c(24.1, 30.8), expand = TRUE) +
    xlab("Longitude")+ ylab("Latitude") +
    theme_classic() + 
    theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA))

rmarkplot2<-  
  ggplot(data=world) + 
  geom_sf(fill="antiquewhite1") +
  geom_sf(data = sites, size = 4, shape = 23, fill = "darkred") +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.63, 26.9), expand=TRUE ) +
    ggtitle ("Location A") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))

rmarkplot3<- 
  ggplot(data=world)+
  geom_sf(fill="antiquewhite1") +
  geom_sf(data = sites, size = 4, shape = 23, fill = "darkred") +
  geom_sf(data = st, fill = NA, colour = "black") + 
  coord_sf(xlim = c(-80.30, -80.00 ), ylim = c(26.62, 26.35), expand=TRUE ) + 
  ggtitle ("Location B") + 
  xlab("Longitude")+ ylab("Latitude") +
  theme_map() + 
  theme(legend.position = "none", panel.grid.major = element_line(colour = gray(.5), linetype = "dashed", size = 0.5),panel.background = element_rect(fill = "aliceblue"), panel.border=element_rect(fill=NA),text=element_text(size=6))
  
  

```


To specify where each plot is displayed with the  `print` function, the arugement `viewport` needs to specify the max width and height of each map, and the minimum x and y coordinates of where the map[s] are located. The `just=` will make a position on how the secondary maps will be displayed. All maps are the same size, until the sizes are adjusted with `width` and `height`.

```{r print_vp}
vp <- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.45, just = c("left","bottom")) 
vp1<- viewport(width = 0.37, height = 0.32, x = 0.22, y =0.15, just = c("left","bottom")) 
```

Now the `print` function can be used the previous spefications that were listed in each plots' `viewport`. 

```{r print_finalmap}

print(rmarkplot)
print(rmarkplot2, vp=vp)
print(rmarkplot3, vp=vp1)


```



# Multiple Maps in one window

Having a way to emphasize locations on a map can also be a graphic that would be useful for biologists. The next steps will be to create blank sqaure plots that will make the map visualization interesting and accurate. In the following map, arrows will point to specific locations.

First we will start with the overall packages needed to build this 

Credit to Ilya Kashnitsky (https://ikashnitsky.github.io/2017/align-six-maps/).

```{r listplot_packages, eval=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(cowplot)
library(ggthemes)
library(gridExtra)
library(grid)
library(lattice)
library(tidyverse)
```


## Map Placement Function 
The function below will allow us to grab plots that we have previously made, with `ggplotGrob` , and create arrows pointing to the maps. 

The `list.plots`, is a simple list command to combine all of the previously created plots.

Afterwards, a function is created to create a custom annotation of each plot, with their corresponding x-min and x-max coordinates. From the bottom left corner of the image, it would be considered 0,0 if it was compared to a mathematical graph. 

The `df.arrows` has the beginning and ending locations of each line. If this case, there are two lines, but more lines can be created by adding more starting locations for `x=` and `y=`. 


```{r listplots_function}

list.plots=list(rmarkplot,rmarkplot2, rmarkplot3)
align_three_plots <- function(list.plots, family = "")
  {
  
gg <- 
  ggplot()+
  coord_equal(xlim = c(0, 80), ylim = c(0, 51), expand=TRUE) +
  annotation_custom(ggplotGrob(rmarkplot), xmin = 0.0, xmax =57, ymin = 0.0, ymax = 50) +
  annotation_custom(ggplotGrob(rmarkplot2),xmin = 58, xmax = 80, ymin = 25, ymax = 51) +
  annotation_custom(ggplotGrob(rmarkplot3), xmin = 58, xmax = 80, ymin = 0.0, ymax = 29) +
  labs(x = NULL, y = NULL) +
  theme_void()


#DF with the coordinates of the 2 arrows
df.arrows <- data.frame(id=1:2,
                        x=c(45.6,45.4),
                        y=c(21.7,19.8),
                        xend=c(58,58),
                        yend=c(35,12))

gg <- 
  gg +
  geom_segment(data = df.arrows %>% filter(id==1),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm"))) +
  geom_segment(data = df.arrows %>% filter(id==2),
             aes(x=x,y=y,xend=xend,yend=yend),
             arrow = arrow(type="closed",length = unit(0.25,"cm")))
}

```


## Creating a Simple Empty Square Plot

The next code will allow up to make a simple empty square plot, that can be cloned so we can place the maps inside. 


```{r listplot_blankplot}
library("ggthemes")

p <- 
  ggplot()+
  theme_map() +
  theme(panel.border = element_rect(color = "black", size = 0.5, fill = NA),aspect.ratio = 1)

```

## Cloning and Final Map 

The function `mget` will clone the square plots three times. The `three` is just to combinge the `align_three_plots` function and the repetition plots in one argument. And the final step is to plot the function and the plots together. 

```{r listplot_finalplot}
library("tidyverse")

plots <- mget(rep("p", 3))

three<- align_three_plots(plots)

plot(three)
```

# Maps an legend placement

The map created below will have a legend placed in the center. First a legend must be create. The map itself with the legend, doesn't have to be accurate, but the legend must be. 

```{r legendmap}

LegendMap<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  scale_fill_viridis(option="plasma", discrete=TRUE)
```

```{r legendmap_maps, include=FALSE}
Map1<- 
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-10.00, 40.00), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

Map2<-
  ggplot(data = world) +
  geom_sf(aes(fill=REGION)) +
  coord_sf(xlim = c(-100.00,-60.00 ), ylim = c(20.00, 50.00), expand=TRUE) +
  scale_fill_viridis(option="plasma", discrete=TRUE) +
  guides(fill=guide_legend(show=FALSE)) +
  theme_map() +
  theme(legend.position= "none", panel.grid.major = element_line(colour = "azure", linetype = "dashed", size = 0.5), panel.background = element_rect(fill = "azure"), panel.border=element_rect(fill=NA),plot.title = element_text(size=16, face="bold"))

```

The function `get_legend`, will grab any legend from any `ggplot` map, or graph. In the function `plot_grid`, the maps that were created,  Map1 and Map2, will be plotted with their associated themes. In this case there is no theme or position, but the option is available. The command `rel_widths`, will specify the size of each pane of the `plot_grid`. The legend will be placed in the middle, and requires the largest size. 


```{r legendmap_prow}

legend <- get_legend(LegendMap+ theme(legend.position="left") )


prow <- plot_grid( Map1 + theme(legend.position="none"), 
                   NULL, Map2 + theme(legend.position="none"),
                   nrow = 1,
                   rel_widths = c(1, 1.5, 1)
)

```

The final map will have the legend "grobbed" and added to the `prow` plot. The first argument is the `legend`, that was taken previously with `get_legend`, and after the first set of restrictions are the left position, the height, and the right position. The positions might have to be tested a few times to check on how accurate the placement is. 

```{r legendmap_final}
prow + draw_grob(legend, 2/4,0/8, 0.003) 
```







# Helpful Resources 

##PDFs for packages

-ggplot2 
https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf
http://ggplot2.tidyverse.org/reference/stat_ecdf.html

-sf
https://cran.r-project.org/web/packages/sf/sf.pdf

-sp
https://cran.r-project.org/web/packages/sp/sp.pdf

-CRAN
https://cran.r-project.org/

-Viridis Color Palettes
https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

-Scale Bar
https://rdrr.io/github/3wen/legendMap/man/scale_bar.html

##Links and various tutorials

-Global Choropleth, world interactive, rotating map

http://ellisp.github.io/blog/2017/06/04/military-gdp


-Ggplotly

http://www.alexejgossmann.com/salaries_by_school_plotly_viz/


-R studio colors

http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf


-Plotting side by side

https://ikashnitsky.github.io/2017/align-six-maps/


-Leaflet

https://cengel.github.io/rspatial/4_Mapping.nb.html

Points with gganimate
http://blog.cultureofinsight.com/2017/06/building-dot-density-maps-with-uk-census-data-in-r/


-Symobls and Points
http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r


-Themes
http://ggplot2.tidyverse.org/reference/theme.html


-Two graphs with same legend

https://github.com/tidyverse/ggplot2/wiki/share-a-legend-between-two-ggplot2-graphs


- Downloading Multiple Zipped files
https://www.r-bloggers.com/batch-downloading-zipped-shapefiles-with-r/

